---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

#### Downloading the required libraries:
```{r warning=FALSE, message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
```

#### Reading the data into R and looking at the first few lines:
```{r warning = FALSE}
setwd("D:/Data Science/Course 5/Week 2/RepData_PeerAssessment1")
unzip("activity.zip")
dat <- read.csv("activity.csv")
head(dat)
```

#### Defining the 'date' column as a date variable:
```{r}
dat$date <- as.POSIXct(strptime(as.character(dat$date),format = "%Y-%m-%d"))
str(dat)
```

## What is mean total number of steps taken per day?

#### Calculating the total number of steps taken per day:
```{r}
tot_step <- dat %>% group_by(date) %>% summarise(total.steps = sum(steps,na.rm = TRUE))
tot_step
```

#### Creating a histogram of total number of steps per day:
```{r message=FALSE, warning=FALSE}
with(tot_step,hist(total.steps,breaks = 10,col="#386cb0",xlab = "Total number of steps taken [per day]",
                   main = "Histogram of total number of steps taken per day"))

dev.copy(png,"Figures/Total number of steps.png", width=480, height=480)
dev.off()
```

#### Calculating the mean and median of total steps per day:
```{r}
mea <- round(mean(tot_step$total.steps),2)
med <- median(tot_step$total.steps)

```

* The mean total number of steps taken per day is: `r mea`
* The median of the total number of steps taken per day is: `r med`

## What is the average daily activity pattern?

#### Calculating the average number of steps per interval accross all days
#### And then generating a time series plot of the average steps per interval:
```{r message=FALSE, warning=FALSE}
ave_dat <- dat %>% group_by(interval) %>% summarise(mean.steps = mean(steps,na.rm = TRUE))
ave_dat <- data.frame(ave_dat)
p <- ggplot(ave_dat,aes(x = interval,y = mean.steps)) +
     geom_line(col = "#1b9e77",size = 1.5) +
     xlab("Interval [min]") +
     ylab("Average number of steps") +
     ggtitle("Time series plot") +
     theme(panel.background = element_blank(),
           panel.border = element_rect(size=0.5,colour = "black",fill = NA),
           panel.grid.major.x = element_blank(),
           panel.grid.major.y = element_blank(),
           axis.text = element_text(size=12,colour="black"),
           axis.title = element_text(size=12,colour = "black"),
           axis.ticks = element_line(size=0.5),
           axis.ticks.length = unit(.15, "cm"),
           plot.title = element_text(face="bold",size = 14, colour = "black",hjust = 0.5))
p

dev.copy(png,"Figures/Average number of steps per interval.png", width=480, height=480)
dev.off()
```

#### Finding the 5 minute interval with the maximum number of steps:
```{r}
max_int <- ave_dat$interval[which(ave_dat$mean.steps==max(ave_dat$mean.steps))]
```

* The interval with the maximum number of steps across all day is: `r max_int`

## Imputing missing values

#### Calculating the total number of missing values in the dataset:
```{r}
num_na <- sum(is.na(dat$steps))
```

* The total number of missing values in the dataset is: `r num_na`

#### Imputing the missing values in data by replacing them with
#### the mean of the 5 minute intervals and assigning new dataset to 'new_dat':
```{r warning=FALSE}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
new_dat <- dat %>% group_by(interval) %>% mutate(steps = impute.mean(steps))
summary(new_dat)
```

#### Calculating the total number of steps taken per day:
```{r}
tot.step <- new_dat %>% group_by(date) %>% summarise(total.steps = sum(steps,na.rm = TRUE))
tot.step
```

#### Creating a histogram of total number of steps per day:
```{r message=FALSE, warning=FALSE}
with(tot.step,hist(total.steps,breaks = 10,col="#f0027f",xlab = "Total number of steps taken [per day]",
                   main = "Histogram of total number of steps taken per day"))

dev.copy(png,"Figures/Total number of steps without NA.png", width=480, height=480)
dev.off()
```

#### Calculating the mean and median of total steps per day:
```{r}
m1 <- round(mean(tot.step$total.steps)/1000,3)
m2 <- round(median(tot.step$total.steps)/1000,3)
```

* The mean total number of steps taken per day is [in thousands]: `r m1`
* The median of the total number of steps taken per day is [in thousands]: `r m2`

## Are there differences in activity patterns between weekdays and weekends?

#### Creating a factor of weekdays versus weekends:
```{r warning=FALSE}
new_dat$day <- character(nrow(new_dat))
new_dat$day[(weekdays(new_dat$date) %in% c("Saturday","Sunday"))] <- "Weekend"
new_dat$day[!(weekdays(new_dat$date) %in% c("Saturday","Sunday"))] <- "Weekday"
new_dat$day <- factor(new_dat$day)
head(new_dat)
```

#### Creating a panel plot of average number of steps per interval
```{r message=FALSE, warning=FALSE}
ave.dat <- new_dat %>%
    group_by(day,interval) %>%
    summarise(mean.steps = mean(steps,na.rm = TRUE))
ave.dat

p <- ggplot(ave.dat,aes(x = interval,y = mean.steps,color = day)) +
     geom_line(size = 1.5) +
     xlab("Interval [min]") +
     ylab("Average number of steps") +
     ggtitle("Time series plot") +
     scale_colour_manual(values = c('#e41a1c','#377eb8')) +
     theme(panel.background = element_blank(),
           panel.border = element_rect(size=0.5,colour = "black",fill = NA),
           panel.grid.major.x = element_blank(),
           panel.grid.major.y = element_blank(),
           axis.text = element_text(size=12,colour="black"),
           axis.title = element_text(size=12,colour = "black"),
           axis.ticks = element_line(size=0.5),
           axis.ticks.length = unit(.15, "cm"),
           plot.title = element_text(face="bold",size = 14, colour = "black",hjust = 0.5),
           legend.position = "n")
p + facet_grid(day ~ .)

dev.copy(png,"Figures/Panel plot.png", width=480, height=480)
dev.off()
```